S-Aligner: Ultra-scalable read mapping on Sunway Taihu Light

### 2017-04-03 s-aligner.tex
initial commit

% REMOVE ME: added for revision
\usepackage{color}

@@

\numberofauthors{7} 
%Xiaohui Duan, Kai Xu, Yuandong Chan, Christian Hundt,\\ Pavan Balaji, Bertil Schmidt, Weiguo Liu\\
\author{
\alignauthor 
Xiaohui Duan\\
    \affaddr{Shandong University}\\
    \affaddr{Jinan City, China}\\
    \email{sunrise.duan@mail.\\sdu.edu.cn}\\
\alignauthor
Kai Xu\titlenote{The first two authors contribute equally to the paper.}\\
    \affaddr{Shandong University}\\
    \affaddr{Jinan City, China}\\
    \email{xu\_kai@mail.sdu.edu.cn}\\
\alignauthor
Yuandong Chan\\
    \affaddr{Shandong University}\\
    \affaddr{Jinan City, China}\\
    \email{yuandong.c@mail.sdu.edu.cn}\\
\and
\alignauthor
Christian Hundt\\
    \affaddr{Johannes Gutenberg University}\\
    \affaddr{Mainz, Germany}\\
    \email{hundt@uni-mainz.de}\\
\alignauthor
Pavan Balaji\\
    \affaddr{Argonne National Laboratory}
    \affaddr{Argonne, Illinois, USA}\\
    \email{balaji@anl.gov}\\
\alignauthor
Bertil Schmidt\\
    \affaddr{Johannes Gutenberg University}\\
    \affaddr{Mainz, Germany}\\
    \email{bertil.schmidt@uni-mainz.de}\\
\and
\alignauthor
Weiguo Liu\\
    \affaddr{Shandong University}\\
    \affaddr{Jinan City, China}\\
    \email{weiguo.liu@sdu.edu.cn}\\
}

@@

The highest level of parallelization employs a coarse-grained partitioning scheme over a block distribution of reads and reference genome using MPI. {\color{red} Weiguo suggests not to mention data size here.} To reduce the excessive data replication of the reference genome index, we employ a partitioning strategy based on a {\em task grid pattern} for the inter-CG parallelization by concurrently assigning pairs of reference genome blocks and read chunks to individual CGs. The two dimensions of the grid are spanned by the reference block identifiers (rows) and read chunk identifiers (columns) as shown in Figure \ref{TaskGrid}(a). Thus, processes in the same row share a unique reference genome (index) block while processes within a column use the same read chunk. Each process executes several cells in one row in order to avoid replication of the relatively big reference index.

@@

We further provide an optional asynchronous data-loading option in case that there is no significantly shorter reference block. In this case we add a row of processes that loads only reads. When other processes in the column are computing alignments, these processes just load new read chunks. After loading and computation have been completed, all processes within a column receive the loaded data by means of an MPI\textunderscore Bcast. In our experiments, this approach  reduces by over 90\% the idle time between computing two read chunks when using 8 processes.  Figure \ref{AsyncRead}(b) illustrates this strategy.

@@

The second level of parallelism exploits the threading capabilities of a CG. Our design within a single node is a comprehensive read mapper based on the described seed-and-extend approach using the three-stage pipeline: filtration, verification, and alignment. The filtration step uses a look-up table of $q$-grams of the assigned reference genome block that can be loaded from a preprocessed index file or alternatively calculated on the fly by a radix sort-based hash table construction method. After the MP performs the lookup of each non-overlapping $q$-gram of a read, it stores the retrieved intervals in a SIMD-friendly manner allowing for efficient vectorization during the subsequent verification step on the SPs. Verification selects all intervals that comply with the restricted edit distance using Myers' bit-parallel algorithm. Afterwards, we employ a banded version of the Smith-Waterman algorithm on the SPs to align the remaining intervals that have passed verification. Figure~\ref{FrmWk} shows the workflow of our intra-CG implementation.

@@

Different approaches have been used in existing aligners: ($i$) BWA \cite{bwa} employs an {\em array of structures} (AoS), ($ii$) RazerS3 \cite{razers3} stores a DNA sequence in a profile using one-hot encoding. While the AoS approach is more space-efficient and cache-friendly, the usage of a profile with one-hot encoding supports bit-parallel computation in a more efficient way.  Because of the small size of the LDM and the requirement of bit parallelism, we employ an encoding strategy that combines an AoS with an SoA ({\em structure of arrays}) approach, which we call an AoSoA ({\em array of structures of arrays}). This mixed strategy builds two bit-vectors in an interleaved fashion. Thus, we can fetch the encoded sequence by one DMA-intrinsic and calculate $\chi(s_{i-1} = s'_{j-1})$ in terms of two logical equality operations. Figure~\ref{MixPack}  illustrates the encoding strategy.

@@

When updating the five bit-vectors according to Equation~\ref{one-hot-myers} along the columns in a bit-parallel fashion, a circular dependency needs to be resolved: $D^0_{i,j}$ depends on $H^-_{i-1, j}$, which in turn depends on $D^0_{i-1,j}$ (a value we have not computed yet). Myers \cite{myers} has shown that this can be solved by using logical operations and an addition. ShenWei's SIMD instructions support 256-bit integer arithmetic, such as $simd\_uaddo\_take\_carry$ (short for SIMD unsigned octa-word add, taking carry) which adds two 256-bit operands and returns a 256-bit unsigned integer. Thus, different from implementations of Myers' algorithm on other architectures (e.g., \cite{chacon}), our implementation does not require additional instructions to process carries generated within a SIMD lane. Unfortunately, there is no 256-bit compare instruction. Thus, we have implemented comparisons in terms of shifting operations.

@@

Since SPs do not have any cache and the latency to access the DDR3 shared memory is high, the usage of the explicitly managed LDM is crucial.  DMA fetching is the most efficient way to transfer data between main memory and LDM (i.e., significantly faster than using functions such as {\em memcpy}). DMA calls are handled by the memory controller, and SPs can continue to perform computation. Thus, we can overlap data transfers from shared memory to LDM and the verification of intervals using Myers algorithm by using asynchronous DMA-fetching intrinsics presented by ShenWei.

@@

Here we use a busy-waiting strategy because the time used for DMA fetching is usually shorter than the time required for verification. Thus, in the common case the DMA reply word needs to be checked just once in order to pass the busy-wait loop. Our experimental results show that our asynchronous data transfer implementation can almost hide the DMA-fetching latency completely and gains a performance improvement of a factor of 22 compared with an implementation based on {\em memcpy}.

@@

The result demonstrates that DMA fetching is key for achieving high performance when using SPs, since a speedup of $\sim$22$\times$ is gained over memcpy. This can be attributed to two factors: (1) DMA transfers require fewer compute resources since they are performed by the memory controller, whereas memcpy uses an SP to write data to memory; (2) DMA fetching can be performed asynchronously, thus enabling the latency of fetching data to be hidden by computation. Furthermore, asynchronous filtration gains $\sim$15\% speedup.

@@

NGS data will continue to grow rapidly in the foreseeable future. Read mapping is a critical and compute-intensive step for a variety of NGS pipelines. While significant efforts have been devoted to optimizing this task, it is still a major bottleneck. In this paper, we have introduced S-Aligner---a highly scalable read mapper specifically designed to fit the characteristics of the Sunway Taihu Light supercomputer and its SW26010 architecture.
It scales almost linearly with over 95\% parallelization efficiency when distributed over more than 50,000 processes. As a result, S-Aligner can map $\approx$ 1.6 TB of read data to the whole human genome in just a few minutes. Moreover, when executed on a single node it can outperform the popular all-mapper RazerS3 executed on eight Xeon E7-8860 CPUs while achieving comparable mapping accuracy.

>>>>

% REMOVE ME: added for revision
% \usepackage{color}

@@

\numberofauthors{7} 
%Xiaohui Duan, Kai Xu, Yuandong Chan, Christian Hundt,\\ Pavan Balaji, Bertil Schmidt, Weiguo Liu\\
\author{
\alignauthor 
%Xiaohui Duan\\
%    \affaddr{Shandong University}\\
%    \affaddr{Jinan City, China}\\
%    \email{sunrise.duan@mail.\\sdu.edu.cn}\\
\alignauthor
%Kai Xu\titlenote{The first two authors contribute equally to the paper.}\\
%    \affaddr{Shandong University}\\
%    \affaddr{Jinan City, China}\\
%    \email{xu\_kai@mail.sdu.edu.cn}\\
\alignauthor
%Yuandong Chan\\
%    \affaddr{Shandong University}\\
%    \affaddr{Jinan City, China}\\
%    \email{yuandong.c@mail.sdu.edu.cn}\\
\and
\alignauthor
%Christian Hundt\\
%    \affaddr{Johannes Gutenberg University}\\
%    \affaddr{Mainz, Germany}\\
%    \email{hundt@uni-mainz.de}\\
\alignauthor
%Pavan Balaji\\
%    \affaddr{Argonne National Laboratory}
%    \affaddr{Argonne, Illinois, USA}\\
%    \email{balaji@anl.gov}\\
\alignauthor
%Bertil Schmidt\\
%    \affaddr{Johannes Gutenberg University}\\
%    \affaddr{Mainz, Germany}\\
%    \email{bertil.schmidt@uni-mainz.de}\\
\and
\alignauthor
%Weiguo Liu\\
%    \affaddr{Shandong University}\\
%    \affaddr{Jinan City, China}\\
%    \email{weiguo.liu@sdu.edu.cn}\\
}

@@

The highest level of parallelization employs a coarse-grained partitioning scheme over a block distribution of reads and reference genome using MPI. In our experiments we process up to 1.6 TB of read data using up to 13,312 SW26010 nodes executing more than 50,000 dispatched processes. Thus, a na\"{\i}ve partitioning of the read file into 50,000 pieces of approximately 32 MB size is not a suitable solution because of the excessive data replication of the reference genome index. Instead, we employ a partitioning strategy based on a {\em task grid pattern} for the inter-CG parallelization by concurrently assigning pairs of reference genome blocks and read chunks to individual CGs. The two dimensions of the grid are spanned by the reference block identifiers (rows) and read chunk identifiers (columns) as shown in Figure \ref{TaskGrid}(a). Thus, processes in the same row share a unique reference genome (index) block while processes within a column use the same read chunk. Each process executes several cells in one row in order to avoid replication of the relatively big reference index.

@@

We further provide an optional asynchronous data-loading strategy in case that there is no significantly shorter reference block. In this case we add a row of processes that loads only reads. When other processes in the column are computing alignments, these processes exclusively load new read chunks. After loading and computation have been completed, all processes within a column receive the loaded data by means of an \texttt{MPI\textunderscore Bcast}. In our experiments, this approach  reduces the idle time between computing two read chunks by over 90\% when using 8 processes. Figure \ref{AsyncRead}(b) illustrates this strategy.

@@

The second level of parallelism exploits the threading capabilities of a CG. Our design within a single node is a comprehensive read mapper based on the described seed-and-extend approach using the three-stage pipeline: filtration, verification, and alignment. The filtration step uses a look-up table of $q$-grams of the assigned reference genome block that can be loaded from a preprocessed index file or alternatively calculated on-the-fly by a radix sort-based hash table construction method. After the MP performs the lookup of each non-overlapping $q$-gram of a read, it stores the retrieved intervals in a SIMD-friendly manner allowing for efficient vectorization during the subsequent verification step on the SPs. Verification selects all intervals that comply with the restricted edit distance using Myers' bit-parallel algorithm. Afterwards, we employ a banded version of the Smith-Waterman algorithm on the SPs to align the remaining intervals that have passed verification. Figure~\ref{FrmWk} shows the workflow of our intra-CG implementation.

@@

Different approaches have been used in existing aligners: BWA \cite{bwa} employs an {\em array of structures} (AoS) while RazerS3 \cite{razers3} stores a DNA sequence in a profile using one-hot encoding. While the AoS approach is more space-efficient and cache-friendly, the usage of a profile with one-hot encoding supports bit-parallel computation in a more efficient way.  Because of the small size of the LDM and the requirement of bit parallelism, we employ an encoding strategy that combines an AoS with an SoA ({\em structure of arrays}) approach, which we call an AoSoA ({\em array of structures of arrays}). This mixed strategy builds two bit-vectors in an interleaved fashion. Thus, we can fetch the encoded sequence by one DMA-intrinsic and calculate $\chi(s_{i-1} = s'_{j-1})$ in terms of two logical equality operations. Figure~\ref{MixPack}  illustrates the encoding strategy.

@@

When updating the five bit-vectors according to Equation~\ref{one-hot-myers} along the columns in a bit-parallel fashion, a circular dependency needs to be resolved: $D^0_{i,j}$ depends on $H^-_{i-1, j}$, which in turn depends on $D^0_{i-1,j}$ (a value we have not computed yet). Myers \cite{myers} has shown that this can be solved by using logical operations and an addition. ShenWei's SIMD instructions support 256-bit integer arithmetic, such as \texttt{simd\_uaddo\_take\_carry} (short for SIMD unsigned octa-word add, taking carry) which adds two 256-bit operands and returns a 256-bit unsigned integer. Thus, different from implementations of Myers' algorithm on other architectures (e.g., \cite{chacon}), our implementation does not require additional instructions to process carries generated within a SIMD lane. Unfortunately, there is no 256-bit compare instruction. Thus, we have implemented comparisons in terms of shifting operations.

@@

Since SPs do not have any cache and the latency to access the DDR3 shared memory is high, the usage of the explicitly managed LDM is crucial.  DMA fetching is the most efficient way to transfer data between main memory and LDM (i.e., significantly faster than using functions such as \texttt{memcpy}). DMA calls are handled by the memory controller, and SPs can continue to perform computation. Thus, we can overlap data transfers from shared memory to LDM and the verification of intervals using Myers' algorithm by using asynchronous DMA-fetching intrinsics presented by ShenWei.

@@

Here we use a busy-waiting strategy because the time used for DMA fetching is usually shorter than the time required for verification. Thus, in the common case the DMA reply word needs to be checked just once in order to pass the busy-wait loop. Our experimental results show that our asynchronous data transfer implementation can almost hide the DMA-fetching latency completely and gains a performance improvement of a factor of 22 compared with an implementation based on \texttt{memcpy}.

@@

The result demonstrates that DMA fetching is key for achieving high performance when using SPs, since a speedup of $\sim$22$\times$ is gained over \texttt{memcpy}. This can be attributed to two factors: (1) DMA transfers require fewer compute resources since they are performed by the memory controller, whereas \texttt{memcpy} uses an SP to write data to memory; (2) DMA fetching can be performed asynchronously, thus enabling the latency of fetching data to be hidden by computation. Furthermore, asynchronous filtration gains $\sim$15\% speedup.

@@

NGS data will continue to grow rapidly in the foreseeable future. Read mapping is a critical and compute-intensive step for a variety of NGS pipelines. While significant efforts have been devoted to optimizing this task, it is still a major bottleneck. In this paper, we have introduced S-Aligner---a highly scalable read mapper specifically designed to fit the characteristics of the Sunway Taihu Light supercomputer and its SW26010 architecture.
It scales almost linearly with over 95\% parallelization efficiency when distributed over more than 50,000 processes. As a result, S-Aligner can map $\approx$ 1.6 TB of read data to the whole human genome in just a few minutes. Moreover, when executed on a single node it can outperform the popular all-mapper RazerS3 executed on eight Xeon E7-8860 CPUs while achieving highly competitive mapping accuracy.
